How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-03-08 doc/paper.tex
seperate main text and supplement references
cleaning supple ref
cleaning up citation formatting

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
%\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[style=numeric-comp,
            backend=biber,
            bibencoding=utf8,
            %maxcitenames=2,
            uniquename=init,
            giveninits=true,
            sorting=none,
            natbib=true,
            url=false,
            doi=false,
            isbn=false]{biblatex}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

@@

@@

\title{How predictable is extinction? Forecasting species survival at million-year timescales}
\author{
  Smits, Peter\\
  %Department of Integrative Biology, University of California Berkeley\\
  \texttt{psmits@berkeley.edu} 
  \and
  Finnegan, Seth\\
  %Department of Integrative Biology, University of California Berkeley\\
  \texttt{sethf@berkeley.edu}
}

@@

@@

@@

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannofossils, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.

@@

\begin{table}[ht]
  \caption{Models and their definitions}
  \begin{threeparttable}
    {
      \def\arraystretch{1.5}
      \begin{tabular}{ l p{3cm} l l }
        Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
        \hline
        C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ age\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}})} \\
        V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) + (1 $|$ age/phylum)} \\ 
        CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff\tnote{g}\phantom{\textsuperscript{g}} +\\temp + temp\_lag\tnote{h}\phantom{\textsuperscript{h}} +\\(1 $|$ age/phylum)} \\
        VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff +\\temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
        \hline
      \end{tabular}
    }
    \begin{tablenotes}
    \item[a] See Equation \ref{eq:model} for full statistical model definition.
    \item[b] Species observation where 1 if time of last observation, otherwise 0.
    \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
    \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
    \item[e] Species are at observation in millions of years.
    \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcaeous nannoplankton).
    \item[g] Change in geographic range since last observation.
    \item[h] Temperature at previous observation.
    \end{tablenotes}
  \end{threeparttable}
  \label{tab:model_def}
\end{table}

@@

We are interested in model adequacy and performance into two contexts: in-sample and out-of-sample predictive performance. ``In-sample'' means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. ``Out-of-sample'' is defined below.

@@

See our code repository https://github.com/psmits/trident for full code details. Our code uses ``tidyverse'' tools such as \texttt{dplyr} \citep{dplyr}, \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}, thus some familiarity with that package ecosystem is necessary to fully comprehend how we've processed our data and results.

@@

\printbibliography

\beginsupplement
\section{Supplementary Material}

@@

>>>>

\usepackage{amsmath, amsthm, amsfonts, amssymb}
\usepackage{microtype, parskip, graphicx}
%\usepackage[comma,numbers,sort&compress]{natbib}
\usepackage[style=numeric-comp,
            backend=biber,
            bibencoding=utf8,
            maxbibnames=10,
            uniquename=init,
            giveninits=true,
            sorting=none,
            natbib=true,
            url=false,
            doi=false,
            isbn=false]{biblatex}
\usepackage[margin=1in]{geometry}
\usepackage{lineno}
\usepackage{longtable}
\usepackage{caption, subcaption, multirow, morefloats, rotating}
\usepackage{wrapfig}
\usepackage{hyperref}
\usepackage{threeparttable}

@@

% https://3d.bk.tudelft.nl/hledoux/blog/fiddling-biblatex/
%-- formatting hell for biblatex
%-- remove "In:"
\renewbibmacro{in:}{}
%-- no "quotes" around titles of chapters/article titles
\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]
{title}{#1}
%-- no punctuation after volume
\DeclareFieldFormat[article]{volume}{ {#1} } 
%-- puts number/issue between brackets
\DeclareFieldFormat[article, inbook, incollection, inproceedings, misc, thesis, unpublished]{number}{\mkbibparens{#1}} 
%-- and then for articles directly the pages w/o any "pages" or "pp." 
\DeclareFieldFormat[article]{pages}{#1}
%-- for some types replace "pages" by "p."
\DeclareFieldFormat[inproceedings, incollection, inbook]{pages}{p. #1}

% https://tex.stackexchange.com/questions/81569/biblatex-parentheses-around-the-volume-number-of-an-article
\renewbibmacro*{volume+number+eid}{%
  \printfield{volume}%
  \printfield{number}%
  \printfield{eid}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}

@@

\title{How predictable is extinction? Forecasting species survival at million-year timescales}
\author{
  Smits, Peter\\
  %Department of Integrative Biology, University of California Berkeley\\
  \texttt{psmits@berkeley.edu} 
  \and
  Finnegan, Seth\\
  %Department of Integrative Biology, University of California Berkeley\\
  \texttt{sethf@berkeley.edu}
}
\date{}

@@

\begin{refsection}

@@

{\bf Keywords:} conservation, paleobiology, extinction

@@

We analyzed microfossil occurrence information which was downloaded from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. All occurrence information was downloaded for calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
This dataset of occurrences was then filtered to just those species which have their first occurrence at most 63 Mya. This choice means that our analysis avoids those taxa which survived the K/Pg boundary, those taxa which arose just after the K/Pg boundary, and means that our occurrence histories line up with the temperature time-series which was used as a predictor of extinction as discussed below.

@@

\begin{table}[ht]
  \caption{Models and their definitions}
  \begin{threeparttable}
    {
      \def\arraystretch{1.5}
      \begin{tabular}{ l p{3cm} l l }
        Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
        \hline
        C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ age\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}})} \\
        V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) + (1 $|$ age/phylum)} \\ 
        CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff\tnote{g}\phantom{\textsuperscript{g}} +\\temp + temp\_lag\tnote{h}\phantom{\textsuperscript{h}} +\\(1 $|$ age/phylum)} \\
        VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff +\\temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
        \hline
      \end{tabular}
    }
    \begin{tablenotes}
    \item[a] See Equation \ref{eq:model} for full statistical model definition.
    \item[b] Species observation where 1 if time of last observation, otherwise 0.
    \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
    \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
    \item[e] Species are at observation in millions of years.
    \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
    \item[g] Change in geographic range since last observation.
    \item[h] Temperature at previous observation.
    \end{tablenotes}
  \end{threeparttable}
  \label{tab:model_def}
\end{table}

@@

We are interested in model adequacy into two contexts: in-sample and out-of-sample predictive performance. In-sample means we are estimating how well our model predicts our observed data given that the model was fit to the entire dataset; this is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. Out-of-sample is defined in context below.

@@

See our code repository at https://github.com/psmits/trident for full code details. The entire analysis was coded in R and uses tidyverse-style tools such as \texttt{dplyr} \citep{dplyr}, and \texttt{purr} \citep{purrr}, and \texttt{tidybayes} \citep{tidybayes}. Additionally, all of our models were written using the \texttt{brms} \citep{brms2017,brms2018} R package, which implements Stan-based Bayesian models which are fit via Hamiltonian Monte Carlo \citep{StanManual}.

@@

\printbibliography
\end{refsection}
\clearpage

\beginsupplement
\begin{refsection}
\section{Supplementary Material}

@@

\printbibliography[title={Supplementary References}]
\end{refsection}
